Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{DISTRIBUTION STATEMENT A. Approved for public release.}

### 2022-03-28 paper/synthesis.tex
latest version of Section 7. Details remain to be checked. Listings may not be up-to-date I fear

Synthesis maps from model and specifications to code. The SPLAT tool
traverses the system architecture looking for occurrences of high
assurance components specified by code contracts; for each such
occurrence it generates a CakeML program. Starting from the semantics
of the code contract described in Section \ref{sec:code-contracts}
a sequence of semantics-preserving steps are taken:
\begin{itemize}
\item $\sem{{\Eqs}\cdot G}$ yields a function on arbitrary depth streams
\item transform to function over depth-1 streams (wellformedness)
\item transform to logic function
\item attach buffer handling to inputs and outputs (contig type parsing)
\item translate to CakeML
\item run CakeML compiler
\end{itemize}

@@

A filter is intended to be simple, although it may make deep semantic
checks. A filter has one input port and one output; messages on the
input that the filter policy admits pass unchanged to the output port;
all others are dropped (not passed on). We have investigated two kinds
of filter. In the first, a relatively shallow scan of the input
suffices to enforce the policy. For example, we have used the
expressive power of Contiguity Types \cite{contiguity-types} to
enforce \emph{lightweight} bounds constraints on GPS coordinates in
UxAS messages. On the other hand, a filter may need to parse the input
buffer into a data structure specified in AGREE and apply a
user-defined \emph{wellformedness} property, also specified in AGREE,
to the data. Arbitrarily complex wellformedness checks can be made in
this way. \figref{fig:filter-spec} shows a combination where the
checking specified by {\small\verb+WELL_FORMED_AUTOMATION_RESPONSE+}
depends on an underlying check specified by the contiguity type
checking bounds on waypoints.

@@

Partiality is an important consideration: steps 2 and 3 above can
fail; the data might not be parseable or the wellformedness
computation could be badly written and fail at runtime. In such cases,
the filter should recover and yield control without passing the input
onwards. In these cases, the filter is behaving as it should, but we
must also guard against situations in which a \emph{correctly
  specified} filter fails at runtime. This kind of defect arises when
the filter \emph{ought} to accept a message, but lack of resources
results in the filter failing to do so. For example, the parse of a
message might need more space than has been allocated; another example
could be if the time slice provided by the scheduler is too short for
the wellformedness computation to finish. Thus resource bounds need to
be included in the correctness argument.

@@

Monitors are intended to track and analyze the externally visible
behavior of system components through time. Therefore, they require
more extensive computational ability than filters. In particular, our
basic notion of a monitor is that it embodies a predicate over its
input and output streams, and is able to access the value of a stream at any
earlier point in time, if necessary. Monitors commonly use state to
keep track of earlier values, unlike filters which, for us, are
typically stateless components. (However, there is nothing in our
approach that forbids stateful filters: they can be realized by
monitors.) A monitor specification is mapped by code generation to a
state transformation function of the following abstract type:
\[
\konst{stepFn} : \mathit{input} \times \mathit{stateVars} \to \mathit{stateVars} \times \mathit{output}
\]

>>>>

Synthesis maps from model and specifications to code. The SPLAT tool
traverses the system architecture looking for occurrences of high
assurance components specified by code contracts; for each such
occurrence it generates a CakeML program. More generally, SPLAT supports
three kinds of behavioral specification from which to generate
code. In increasing order of expressiveness, these are:

\paragraph{Regular expressions.} Often the well-formedness of data on connections
can be enforced by matching against a regular expression. In other
words, the data can be characterized by a \emph{regular language}. As
is well known, regular expressions can be translated to efficient
deterministic finite state machine (DFA)
implementations. In \cite{formal-filter-synth-langsec,case-verified-filter}
we report on the creation and application of filters from a verified
regular expression-to-DFA compiler.

\paragraph{Contiguity types\cite{contiguity-types}.}
A contiguity type, like regular expressions, is effective at
describing data at the network message level, \eg, as flat
strings. The notation is however more expressive, being able to
declare many kinds of so-called \emph{self-describing} messages, \ie,
those where message structure is dependent on data held elsewhere in
the message. Figure \ref{fig:filter-spec} gives a contiguity type
specification for the welll-formed waypoints of our example system. We
have formalized and verified a contiguity type matcher. The matcher
code has been used in our case studies.

\paragraph{AGREE code contracts.} When full computational power is needed, we can use the
expressive power of AGREE code contracts to write arbitrarily complex
filter or monitor operations, as we have already seen. Our formal
model of AGREE syntax and semantics is applied to extract code
contracts and map them to equivalent logic functions.

\noindent For each of these three approaches, we take the following steps to
arrive at an executable.

\begin{enumerate}
\item Map from specification to computable function in logic, producing
a proof of equivalence.
\item Add parsers and printers (specified by contiguity types) for input and output.
\item Map to CakeML and add calls to input/output library routines via
      the CakeML foreign function interface.
\item Invoke the CakeML compiler.
\end{enumerate}

%% Starting from the semantics of the code contract described in
%% Section \ref{sec:code-contracts} a sequence of semantics-preserving
%% steps are taken:
%% \begin{itemize}
%% \item $\sem{{\Eqs}\cdot G}$ yields a function on arbitrary depth streams
%% \item transform to Pascalish
%% \item transform to logic function
%% \item attach buffer handling to inputs and outputs (contig type parsing)
%% \item translate to CakeML
%% \item run CakeML compiler
%% \end{itemize}

@@

A filter is intended to be simple, although it may make deep semantic
checks. A filter has one input port and one output; messages on the
input that the filter policy admits pass unchanged to the output port;
all others are dropped (not passed on). We have investigated two kinds
of filter. In the first, a relatively shallow scan of the input
suffices to enforce the policy. For example, we have used the
expressive power of regular expressions and Contiguity
Types \cite{contiguity-types} to enforce \emph{lightweight} bounds
constraints on GPS coordinates in UxAS messages. On the other hand, a
filter may need to parse the input buffer into a data structure
specified in AGREE and apply a user-defined \emph{wellformedness}
property, also specified in AGREE, to the data. Arbitrarily complex
wellformedness checks can be made in this
way. \figref{fig:filter-spec} shows a combination where the checking
specified by {\small\verb+WELL_FORMED_AUTOMATION_RESPONSE+} depends on
an underlying check specified by the contiguity type checking bounds
on waypoints.

@@

Partiality is an important consideration: steps 2 and 3 above can
fail; the data might not be parseable or the wellformedness
computation could be badly written and fail at runtime. In such cases,
the filter should recover and yield control without passing the input
onwards. In these cases, the filter is behaving as it should, but we
must also guard against situations in which a \emph{correctly
specified} filter fails at runtime. This kind of defect arises when
the filter \emph{ought} to accept a message, but lack of resources
results in the filter failing to do so. For example, the parse of a
message might need more space than has been allocated; another example
could be if the time slice provided by the scheduler is too short for
the wellformedness computation to finish. Thus resource bounds need to
be included in the correctness argument. Some preliminary work on this
has been done in CakeML \cite{cakeml-space-cost}.

@@

Monitors are intended to track and analyze the externally visible
behavior of system components through time. Therefore, they tend to
require more extensive computational ability than filters. In
particular, our basic notion of a monitor is that it embodies a
relation over its input and output streams, and is able to access the
value of a stream at any earlier point in time, if necessary. Monitors
commonly use state to keep track of earlier values, unlike filters
which, for us, are typically stateless components. (However, there is
nothing in our approach that forbids stateful filters: they can be
realized by monitors.) A monitor specification is mapped by code
generation to a state transformation function of the following
abstract type:
\[
\konst{stepFn} : \mathit{input} \times \mathit{stateVars} \to \mathit{stateVars} \times \mathit{output}
\]
